# Trends

## Least squares estimation for linear regression trend

We begin by taking the partial derivatives with respect to $\beta_0$.

\[
\frac{\partial}{\partial{\beta_0}} \mathcal{Q}(\beta_0, \beta_1) =
  -2\sum_{t=1}^n (Y_t - \beta_0 - \beta_1 t)
\]

We set it to $0$ and from this retrieve

\begin{align*}
-2\sum_{t=1}^n (Y_t - \beta_0 - \beta_1 t) = & 0 \implies \\
\sum_{t=1}^n Y_t - n\beta_0 - \beta_1 \sum_{t=1}^n t = & 0 \implies \\
\beta_0 = \frac{\sum_{t=1}^n Y_t - \beta_1 \sum_{t=1}^n t}{n} = &
  \bar{Y} - \beta_1 \bar{t}
\end{align*}

Next, we take the partial derivative with respect to $\beta_1$;

\[
\frac{\partial}{\partial{\beta_1}} \mathcal{Q}(\beta_0, \beta_1) =
  -2\sum_{t=1}^n t(Y_t - \beta_0 - \beta_1 t)
\]

Setting this to $0$ as well, multiplying both sides with $-1/2$ and rearranging
results in

\begin{align*}
-2\sum_{t=1}^n t (Y_t - \beta_0 - \beta_1 t) = & 0 \implies \\
\beta_1 \sum_{t=1}^n t^2 = & \sum_{t=1}^n Y_t t - \beta_0 \sum_{t=1}^n t
\end{align*}

Then, substituting with the result gained previously for $\beta_0$, we get

\begin{align*}
\beta_1 \sum_{t=1}^n t^2 = & \sum_{t=1}^n Y_t t - 
 \left( \frac{\sum_{t=1}^n Y_t}{n} - \beta_1 \frac{\sum_{t=1}^n}{n} \right)
 \sum_{t=1}^n t \iff \\
\beta_1 \left( \sum_{t=1}^n t^2 - \frac{(\sum_{t=1}^n t)^2}{n} \right) = & 
  \sum_{t=1}^n Y_t t - \frac{\sum_{t=1}^n Y_t \sum_{t=1}^n t}{n} \iff \\
\beta_1 = & \frac{n\sum_{t=1}^n Y_tt - \sum_{t=1}^nY_t \sum_{t=1}^n t}{n \sum_{t=1}^n t^2 - \left( \sum_{t=1}^n t \right)^2} = 
  \frac{\sum_{t=1}^n (Y_t - \bar{Y})(t-\bar{t})}{\sum_{t=1}^n (t-\bar{t})^2} \quad \square
\end{align*}


## Variance of mean estimator

\[
  \bar{Y} = \frac{1}{n}\sum_{t=1}^n Y_t = \frac{1}{n} \sum_{t=1}^n(\mu + e_t - e_{t-1}) = 
    \mu + \frac{1}{n} \sum_{t=1}^n (e_t - e_{t-1}) = \mu + \frac{1}{n}(e_n - e_0)
\]

\[
  \text{Var}[\bar{Y}] = \text{Var}[\mu + \frac{1}{n}(e_n - e_0)] =
    \frac{1}{n^2}(\sigma_e^2 + \sigma_e^2) = \frac{2\sigma_e^2}{n^2}
\]

It is uncommon for the sample size to have such a large impact on the variance
estimator for the sample mean.

Setting $Y_t = \mu + e_t$ instead gives

\[
  \bar{Y} = \frac{1}{n}\sum_{t=1}^n Y_t = \frac{1}{n} \sum_{t=1}^n(\mu + e_t) = 
    \mu + \frac{1}{n} \sum_{t=1}^n e_t
\]

\[
  \text{Var}[\bar{Y}] = \text{Var} \left[ \mu + \frac{1}{n} \sum_{t=1}^n e_t \right] =
    0 + \frac{1}{n^2} \times n \sigma_e^2 = \frac{\sigma_e^2}{n}.
\]

## Variance of mean estimator #2

\[
  \bar{Y} = \frac{1}{n} \sum_{t=1}^n(\mu + e_t + e_{t-1}) = 
    \mu + \frac{1}{n} \sum_{t=1}^n (e_t + e_{t-1}) = \mu + \frac{1}{n} \left( e_n + e_0 + 2 \sum_{t=1}^{n-1} t \right)
\]

\[
  \text{Var}[\bar{Y}] = \frac{1}{n^2}(\sigma_e^2 + \sigma_e^2 + 4(n-1) \sigma_e^2 ) = \frac{1}{n^2}2(2n-1)\sigma_e^2
\]

Setting $Y_t = \mu + e_t$ instead gives the result from 3.2. We note that for
large $n$ the variance if approximately four times larger with
$Y_t = \mu + e_t + e_{t-1}$.

## Hours

### a {-}

```{r, fig.cap="Monthly values of the average hours worked per week in the U.S. manufacturing sector."}
library(TSA)
data("hours")
xyplot(hours)
```

In Figure 1 we see a steep incline between 83 and 84. There also appears to be a seasonal
trend with generally longer work hours later in the year apart from the summer;
1984, however, does not exhibit as clear a pattern.

### b {-}

```{r, fig.cap="Monthly values of average hours worked per week with superposed initials of months."}
months <- c("J", "A", "S", "O", "N", "D", "J", "F", "M", "A", "M", "J")

xyplot(hours, panel = function(x, y, ...) {
  panel.xyplot(x, y, ...)
  panel.text(x = x, y = y, labels = months)
})
```

Here, in Figure 2, our interpretation is largely the same. It is clear that
December stands out as the month with the longest weekly work hours whilst
February and January are low-points, demonstrating a clear trend.

## Wages

### a {-}

```{r, fig.cap="Monthly average hourly wages for workers in the U.S. apparel and textile industry."}
data("wages")
xyplot(wages, panel = function(x, y, ...) {
  panel.xyplot(x, y, ...)
  panel.text(x, y, labels = months)
})
```

There is a positive trend with seasonality: August is a low-point for wages. 
Generally, there seems to be larger increases in the fall.

### b {-}

```{r}
wages_fit1 <- lm(wages ~ time(wages))
summary(wages_fit1)
wages_rst <- rstudent(wages_fit1)
```

### c {-}

```{r wages_resid, fig.cap="Residual plot"}
xyplot(wages_rst ~ time(wages_rst), type = "l",
       xlab = "Time", ylab = "Studentized residuals")
```

We still seem to have autocorrelation related to the time and not white noise.

### d {-}

```{r}
wages_fit2 <- lm(wages ~ time(wages) + I(time(wages)^2))
summary(wages_fit2)
wages_rst2 <- rstudent(wages_fit2)
```

### e {-}

```{r wages_quad_resid, fig.cap="Residual plot for our quadratic model."}
xyplot(wages_rst2 ~ time(wages_rst), type = "l",
       xlab = "Time", ylab = "Studentized residuals")
```

This looks more like random noise but there is still clear autocorrelation 
between the fitted residuals that we have yet to capture in our model.

## Beer sales

### a {-}

```{r, fig.cap = "Monthly U.S. beer sales."}
data(beersales)
xyplot(beersales)
```

Clear seasonal trends. There is an initial positive trend from 1975 to around
1981 that then levels out.

### b {-}

```{r, fig.cap = "Monthly U.S. beer sales annotated with the months' initials."}
months <- c("J", "F", "M", "A", "M", "J", "J", "A", "S", "O", "N", "D")

xyplot(beersales,
       panel = function(x, y, ...) {
         panel.xyplot(x, y, ...)
         panel.text(x, y, labels = months)
       })
```

It is now evident that the peaks are in the warm months and the slump in the
winter and fall months. December is a particular low point, while May, June,
and July seem to be the high points.

### c {-}

```{r}
beer_fit1 <- lm(beersales ~ season(beersales))
summary(beer_fit1)
```

All comparisons are made against january. The model helpfully explains 
approximately 0.71 of the variance and is statistically significant. Most 
of the factors are significant (mostly the winter months as expected).

### d {-}

```{r, fig.cap="Beer sales residual plot."}
xyplot(rstudent(beer_fit1) ~ time(beersales), type = "l",
       xlab = "Fitted values", ylab = "Time",
       panel = function(x, y, ...) {
         panel.xyplot(x, y, ...)
         panel.text(x, y, labels = season(beersales))
       })
```

### e {-}








