# Models for stationary time series

## First principles

We have the process
\[
Y_t = 5  + e_t - \frac{1}{2}e_{t-i} + \frac{1}{4}e_{t-2}
\]
and begin by working out its variance
\[
\begin{aligned}
\text{Var}(Y_t) & = \text{Var}(5  + e_t - \frac{1}{2}e_{t-i} + \frac{1}{4}e_{t-2})\\
                & = \text{Var}(e_t) + \frac{1}{4}\text{Var}(e_t) + \frac{1}{16}\text{Var}(e_t)\\
                & = \frac{21}{16}\sigma_e^2
\end{aligned}
\]
and then the autocovariance at lag 1
\[
\begin{aligned}
\text{Cov}(Y_t, Y_{t-1}) & = \text{Cov}(5+e_t-\frac{1}{2}e_{t-1}+\frac{1}{4}e_{t-2}, 5+e_{t-1}-\frac{1}{2}e_{t-2}+\frac{1}{4}e_{t-3})\\
                         & = \text{Cov}(-\frac{1}{2}e_{t-1},e_{t-1}) + \text{Cov}(\frac{1}{4}e_{t-2},-\frac{1}{2}e_{t-2}) \\
                         & = -\frac{1}{2}\text{Var}(e_{t-1}) -\frac{1}{8}\text{Var}(e_{t-2})\\
                         & = -\frac{5}{8}\sigma_e^2
\end{aligned}
\]
lag 2
\[
\begin{aligned}
\text{Cov}(Y_t, Y_{t-2}) & = \text{Cov}(5+e_t-\frac{1}{2}e_{t-1}+\frac{1}{4}e_{t-2}, 5+e_{t-2}-\frac{1}{2}e_{t-3}+\frac{1}{4}e_{t-4})\\
                         & = \frac{1}{4}\text{Var}(e_{t-2}) \\
                         & = \frac{1}{4}\sigma_e^2
\end{aligned}
\]
and lag 3
\[
\text{Cov}(Y_t, Y_{t-2}) = \text{Cov}(5+e_t-\frac{1}{2}e_{t-1}+\frac{1}{4}e_{t-2}, 5+e_{t-2}-\frac{1}{2}e_{t-3}+\frac{1}{4}e_{t-4}) = 0
\]
which results in the autocorrelation
\[
\rho_k = \begin{cases}
           1 & k = 0\\
           \frac{-\frac{5}{8}\sigma_e^2}{\frac{21}{16}\sigma_e^2}=-\frac{10}{21} & k = 1\\
           \frac{\frac{1}{4}\sigma_e^2}{\frac{21}{16}\sigma_e^2}=\frac{4}{21} & k = 2\\
           0 & k = 3\\
         \end{cases} \tag*{$\square$}
\]

## Sketch autocorrelations

### a {-}
```{r, fig.cap = "Autocorrelation with $\\theta_1 = 0.5$ and $\\theta_2 = 0.4$"}
tacf(ma = list(-0.5, -0.4))
```

### b {-}

```{r, fig.cap = "Autocorrelation with $\\theta_1 = 1.2$ and $\\theta_2 = -0.7$"}
tacf(ma = list(-1.2, 0.7))
```

### c {-}

```{r, fig.cap = "Autocorrelation with $\\theta_1 = -1$ and $\\theta_2 = -0.6$"}
tacf(ma = list(1, 0.6))
```

## Max and min correlations for MA(1)

For
\[
\rho_1 = \frac{-\theta}{1+\theta^2}
\]
we retrieve extreme values at 
\[
\frac{\partial}{\partial \theta}\rho_1 = \frac{-1(1+\theta^2)-2\theta(-\theta)}{(1+\theta^2)^2} = \frac{\theta^2 - 1}{(1+\theta^2)^2} = 0
\]
when $t = \begin{cases}-1\\1\end{cases}$, which gives us
\[
\begin{aligned}
\max \rho_1 & = \frac{-1(-1)}{1+(-1)^2} = 0.5\\
\min \rho_1 & = \frac{-1}{1+1^2} = -0.5
\end{aligned}
\]
which we graph in figure \@ref(fig:minmax)

```{r minmax, fig.cap = "Autocorrelation at lag one for MA(1) with max and min annotated."}
theta <- seq(-10, 10, by = 0.01)
p1 <- (-theta) / (1 + theta^2)
plot(theta, p1, type = "l")
points(theta[which.max(p1)], max(p1))
points(theta[which.min(p1)], min(p1))
```

## Non-uniqueness of MA(1)

\[
\frac{-\frac{1}{\theta}}{1 + \left( \frac{1}{\theta}\right)^2} = \frac{-\frac{1}{\theta}\times\theta^2}{\left( 1 + \frac{1}{\theta^2} \right) \theta^2} = \frac{-\theta}{1+\theta^2} \tag*{$\square$}
\]

## Sketch more autocorrelations

```{r autocorrelation, fig.width = 2.2, fig.height = 2, fig.show = "hold", fig.cap = "ACF for various AR(1) processes."}
theta <- c(0.6, -0.6, 0.95, 0.3)
lag <- c(10, 10, 20, 10)
for (i in seq_along(theta)) {
  print(tacf(ar = theta[i], lag.max = lag[i]))
}
```

## Difference function for AR(1)

### a {-}

\[
\begin{aligned}
\text{Cov}(\triangledown Y_t, \triangledown Y_{t-k}) & = \text{Cov}(Y_t-Y_{t-1}, Y{t-k}-Y_{t-k-1})\\
                                                     & = \text{Cov}(Y_t, Y_{t-k}) - \text{Cov}(Y_{t-1},Y_{t-k}) - \text{Cov}(Y_t, Y_{t-k-1}) + \text{Cov}(Y_{t-1}, Y_{t-k-1})\\
                                                     & = \frac{\sigma_e^2}{1-\phi^2}(\phi^2 - \phi^{k-1}-\phi^{k+1}+\phi^k) \\
                                                     & = \frac{\sigma_e^2}{1-\phi^2}\phi^{k-1}(2\phi-\phi2-1)\\
                                                     & = - \frac{\sigma_e^2}{1-\phi^2}(1-\phi)^2\phi^{k-1}\\
                                                     & = - \sigma_e^2 \frac{(1-\phi)^2}{(1-\phi)(1+\phi)}\\
                                                     & = -\sigma_e^2 \frac{1-\phi}{1+\phi}\phi^{k-1}
\end{aligned}
\]
as required.

### b {-}

\[
\begin{aligned}
  \text{Var}(W_t) & = \text{Var}(Y_t-Y_{t-1})\\
                  & = \text{Var}(\phi_1Y_{t-1}+e_t-Y_{t-1})\\
                  & = \text{Var}(Y_{t-1}(\phi-1)+\sigma_e^2)\\
                  & = (\phi-1)^2\text{Var}(Y_{t-1}) + \text{Var}(e_t)\\
                  & = \frac{\sigma_e^2}{1-\phi^2}(\phi^2-2\phi+1) + \sigma_e^2\\
                  & = \frac{\sigma_e^2(\phi^2-2\phi+1+1-\phi^2)}{1-\phi^2}\\
                  & = \frac{2\sigma_e^2(1-\phi)}{1-\phi^2} \\
                  & = \frac{2\sigma_e^2}{1+\phi} \tag*{$\square$}
\end{aligned}
\]

## Characteristics of several models

### a {-}

Only correlation at lag 1.

### b {-}

Only autocorrelation at lag 1 and 2. Shape of process depends on values of
coefficients.

### c {-}

Exponentially decaying correlation from lag 0.

### d {-}

Different patterns in ACF that depends on whether roots are complex or real.

### e {-}

Exponentially decaying correlations from lag 1.


## AR(2)

First, we have variance
\[
\text{Var}(Y_t) = \text{Var}(\phi_2 Y_{t-2} + e_t) = \phi_2^2 \text{Var}Y_{t-2} + \sigma_e^2
\]
which, assuming stationarity, is equivalent to
\[
\text{Var}(Y_{t-2}) = \phi_2^2\text{Var}(Y_{t-2}) + \sigma_e^2 \iff \\
\sigma_e^2 = (1-\phi_2^2) \text{Var}(Y_{t-2}) \iff \\
\text{Var}(Y_{t-2}) = \frac{\sigma_e^2}{1-\phi^2_2}
\]
which requires that $-1 < \phi_2 < 1$ since $\text{Var}(Y_{t-2}) \geq 0$.

## Sketching AR(2) processes

### a {-}

\[
\begin{split}
\rho_1 & = 0.6\rho_0 + 0.3\rho{-1} = 0.6 + 0.3\rho_1 = 0.8571 \\
\rho_2 & = 0.6\rho_1+0.3\rho_0 = 0.81426\\
\rho_3 & = 0.6\rho_2 + 0.3\rho_1 = 0.7457
\end{split}
\]

The roots to the characteristic equation are given by

\[
\frac{\phi_1 \pm \sqrt{\phi_1^2 + 4\phi_2}}{-2\phi_2} = \frac{0.6 \pm \sqrt{0.6 + 4 \times 0.3}}{-2 \times 0.3} = -1 \pm 2.0817 = \{1.0817, -3.0817\}.
\]

Since both of these roots exceed 1 in absolute value, they are real. Next,
we sketch the theoretical autocorrelation function (\@ref(fig:ar2a)).

```{r ar2a, fig.cap = "ACF for AR(2) with $\\phi_1 = 0.6, \\phi_2 = 0.3$."}
tacf(ar = c(0.6, 0.3))
```

### b {-}

Next we write a function to do the work for us.

```{r}
ar2solver <- function(phi1, phi2) {
  roots <- polyroot(c(1, -phi1, -phi2))
  cat("Roots:\t\t", roots, "\n")
  
  if (any(Im(roots) > sqrt(.Machine$double.eps))) {
    damp <- sqrt(-phi2)
    freq <- acos(phi1 / (2 * damp))
    
    cat("Dampening:\t", damp, "\n")
    cat("Frequency:\t", freq, "\n")
  }
  
  tacf(ar = c(phi1, phi2))
}
```

```{r, fig.cap = "ACF for AR(2) with $\\phi_1 = -0.4, \\phi_2 = 0.5$."}
ar2solver(-0.4, 0.5)
```

### c {-}

```{r, fig.cap = "ACF for AR(2) with $\\phi_1 = 1.2, \\phi_2 = -0.7$."}
ar2solver(1.2, -0.7)
```

### d {-}

```{r, fig.cap = "ACF for AR(2) with $\\phi_1 = -1, \\phi_2 = -0.6$."}
ar2solver(-1, -0.6)
```

### e {-}

```{r, fig.cap = "ACF for AR(2) with $\\phi_1 = 0.5, \\phi_2 = -0.9$."}
ar2solver(0.5, -0.9)
```

### f {-}

```{r, fig.cap = "ACF for AR(2) with $\\phi_1 = -0.5, \\phi_2 = -0.6$."}
ar2solver(-0.5, -0.6)
```

## Sketch ARMA(1,1) models

### a {-}

```{r, fig.cap = "ACF for ARMA(1,1) with $\\phi = 0.7$ and $\\theta = 0.4$."}
tacf(ar = 0.7, ma = -0.4)
```

### b {-}

```{r, fig.cap = "ACF for ARMA(1,1) with $\\phi = 0.7$ and $\\theta = -0.4$."}
tacf(ar = 0.7, ma = 0.4)
```

## ARMA(1,2)

### a {-}

\[
\begin{aligned}
\text{Cov}(Y_t, Y_{t-k}) & = \text{E}[(0.8Y_{t-1}+e_t+0.7e_{t-1}+0.6e_{t-2})Y_{t-k}] - \text{E}(Y_t)\text{E}(Y_{t-k})\\
                         & = \text{E}(0.8Y_{t-1}Y_{t-k} + Y_{t-k}e_t + 0.7e_{t-1}Y_{t-k}+0.6e_{t-2}Y_{t-k}) - 0\\
                         & = 0.8\text{E}(Y_{t-1}Y_{t-k}) + \text{E}(Y_{t-k}e_t) + 0.7\text{E}(e_{t-1}Y_{t-k}) + 0.7\text{E}(e_{t-2}Y_{t-k})\\
                         & = 0.8\text{E}(Y_{t-1}Y_{t-k})\\
                         & = 0.8\text{Cov}(Y_t,Y_{t-k+1})\\
                         & = 0.8\gamma_{k-1} & \square
\end{aligned}
\]

### b {-}

\[
\begin{split}
\text{Cov}(Y_t,Y_{t-2}) & = \text{E}[0.8Y_{t-1}+e_t+0.7e_{t-1}+0.6e_{t-2})Y_{t-2}]\\
                        & = \text{E}[(0.8Y_{t-1}+0.6e_{t-2})Y_{t-2}]\\
                        & = 0.8\text{Cov}(Y_{t-1},Y_{t-2})+0.6\text{E}(e_{t-2}Y_{t-2})\\
                        & = 0.8\gamma_1 + 0.6\text{E}(e_t Y_t)\\
                        & = 0.8\gamma_1 + 0.6\text{E}[e_t(0.8Y_{t-1}+e_t+0.7e_{t-1}+0.6e_{t-2})]\\
                        & = 0.8\gamma_1 + 0.6\sigma_e^2 \iff \\
\rho_2                  & = 0.8\rho_1+0.6\sigma_e^2/\gamma_0 & \square
\end{split}
\]

## Two MA(2) processes

### a {-}

For $\theta_1 = \theta_2 = 1/6$ we have
\[
\rho_k = \frac{-\frac{1}{6}+\frac{1}{6}\times\frac{1}{6}}{1 + \left(\frac{1}{6}\right)^2 + \left(\frac{1}{6}\right)^2} = 
\frac{\frac{1}{6}\left(\frac{1}{6}-1\right)}{1 + \frac{2}{36}} = - \frac{5}{38}.
\]

For $\theta_1 = -1$ and $\theta_2 = 6$,
\[
\rho_k = \frac{1-6}{1+1^2+36} = - \frac{5}{38} \tag*{$\square$}.
\]

### b {-}

For $\theta_1 = \theta_2 = 1/6$ we have roots given by 
\[
\frac{\frac{1}{6} \pm \sqrt{\frac{1}{36}+ 4 \times \frac{1}{6}}}{-2\times \frac{1}{6}} =
- \frac{1}{2} \pm \frac{\sqrt{\frac{25}{36}}}{-\frac{1}{3}} = - \frac{1}{2} \pm \frac{\frac{5}{6}}{\frac{1}{3}} = \{-3, -2\}
\]

and for $\theta_1 = -1$ and $\theta_2 = 6$,
\[
\frac{-1 \pm \sqrt{1 + 4 \times 6}}{-2\times6} = \frac{-1\pm 5}{-12} = \frac{1}{12} \pm \frac{5}{12} = \left\{-\frac{1}{3}, \frac{1}{2}\right\}
\]

## Autocorrelation in MA(1)

\[
\begin{aligned}
& \text{Var}(Y_{n+1}+Y_n+Y_{n-1}+ \dots + Y_1)  = \left((n+1)+2n\rho_1\right)\gamma_0 = \left(1+n(1+2\rho_1)\right)\gamma_0\\
& \text{Var}(Y_{n+1}-Y_n+Y_{n-1}- \dots + Y_1)  = \left((n+1)-2n\rho_1 \right)\gamma_0 = \left(1 + n(1-2\rho_1)\right)\gamma_0
\end{aligned}
\]

\[
\begin{cases}
\left(1+n(1+2\rho-1)\right) \geq 0 \\
\left(1+n(1-2\rho-1)\right) \geq 0 \\
\end{cases} \iff

\begin{cases}
1+n+2\rho_1n \geq 0 \\
1+n-2\rho_1n \geq 0 \\
\end{cases} \iff

\begin{cases}
\rho_1 \geq \frac{-(n+1)}{2n}\\
\rho_1 \leq \frac{n+1}{2n}
\end{cases} \iff

\begin{cases}
\rho_1 \geq -\frac{1}{2}(1+\frac{1}{n})\\
\rho_1 \leq \frac{1}{2}(1+\frac{1}{n})
\end{cases}
\]
where $\rho_1 \geq |1/2|$ for all $n$.

## Zero-mean stationary process

We set $Y_t=e_t−θe_t−1$ and then we have

$$
\begin{split}
e_t & = \sum_{j=0}^\infty \theta^j Y_{t-j}\quad \text{and expanding into} \\
    & = \sum_{j=1}^\infty \theta^j Y_{t-j} + \theta^0 Y_{t-0} \\
    & \iff \\
Y_t & = e_t - \sum_{j=1}^\infty \theta^j Y_{t-j}
\end{split}
$$
which is equivalent to
\[
Y_t = \mu_0 + (1 + \theta B + \theta^2 B^2 + \dots + \theta^n B^n)e_t
\]
which is the definition of a MA(1) process where $B$ is the backshift operator
such that $Y_t B^k =Y_{t-k}$.

## Stationarity prerequisite of AR(1)

\[
\begin{split}
\text{Var}(Y_t) & = \text{Var}(\phi Y_{t-1}+e_t) = \phi^2\text{Var}(Y_{t-1})+\sigma_e^2\\
                & = \phi^2 \text{Var}(\phi Y_{t-2}+e_t)+\sigma_e^2\\
                & = \phi^4 \text{Var}(Y_{t-2})+2\sigma_e^2\\
                & = \phi^{2n}\text{Var}(Y_{t-n})+n\sigma_e^2
\end{split}
\]
And $\lim_{n \rightarrow \infty} \text{Var}(Y_t) = \infty$ if $|\phi$=1$, which
is impossible.

## Nonstationary AR(1) process

### a {-}

\[
\begin{split}
Y_t & = \phi Y_{t-1}+e_t \implies \\
- \sum_{j=1}^\infty \left(\frac{1}{3}\right)^j e_{t+j}     & = 3 \left(-\sum_{j=1}^\infty \left(\frac{1}{3}\right)^j e_{t-1+j}\right) + e_t\\
- \sum_{j=1}^\infty \left(\frac{1}{3}\right)^{j+1} e_{t+j} & = -\sum_{j=1}^\infty \left(\frac{1}{3}\right)^j e_{t-1+j} + \frac{1}{3} e_t\\
- \sum_{j=1}^\infty \left(\frac{1}{3}\right)^{j+1} e_{t+j} & = -\sum_{j=2}^\infty \left(\frac{1}{3}\right)^j e_{t-1+j}\\
- \sum_{j=1}^\infty \left(\frac{1}{3}\right)^{j+1} e_{t+j} & = -\sum_{j+1=2}^\infty \left(\frac{1}{3}\right)^{j+1} e_{t+j} & \square
\end{split}
\]

### b {-}

\[
\text{E}(Y_t) = \text{E}(\sum_{j=1}^\infty \left(\frac{1}{3}\right)^j e_{t+j}) = 0
\]
since all terms are uncorrelated white noise.

\[
\begin{gathered}
\text{Cov}(Y_t,Y_{t-1}) = \text{Cov}\left( -\sum_{j=1}^\infty \left(\frac{1}{3}\right)^j e_{t+j}, \sum_{j=1}^\infty \left(\frac{1}{3}\right)^j e_{t+j-1} \right) = \\
\text{Cov}\left(-\frac{1}{3}e_{t+1} - \left(\frac{1}{3}\right)^2e_{t+2} - \dots - \left(\frac{1}{3}\right)^n e_{t+n},
                -\frac{1}{3}e_{t} - \left(\frac{1}{3}\right)^2e_{t+1} - \dots - \left(\frac{1}{3}\right)^n e_{t+n-1} \right) = \\
\text{Cov}\left(-\frac{1}{3}e_{t+1},-\frac{1}{3^2}e_{t+1}\right) +
  \text{Cov}\left(-\frac{1}{3}e_{t+2},-\frac{1}{3^3}e_{t+2}\right) + \dots +
  \text{Cov}\left(-\frac{1}{3}e_{t+n},-\frac{1}{3^{n+1}}e_{t+n}\right) = \\
\frac{1}{26}\sigma_e^2\left(1 + \frac{1}{3} + \frac{1}{3^2} + \dots + \frac{1}{3^n} \right)
\end{gathered}
\]
which is free of $t$.

### c {-}

It is unsatisfactory because Y_t depends on future observations.

## Solution to AR(1)

### a {-}

\[
\frac{1}{2}\left(10\left(\frac{1}{2}\right)^{t-1} + e_{t-1} + \frac{1}{2}e_{t-2} + \left(\frac{1}{2}\right)^2e_{t-3} + \dots +
  \left(\frac{1}{2}\right)^{n-1}e_{t-n}\right) + e_{t-1} = \\
10 \left(\frac{1}{2}\right)^t + \frac{1}{2}e_{t-1} + \left(\frac{1}{2}\right)^2e_{t-2} + \left(\frac{1}{2}\right)^3 e_{t-3} +
  \dots + \left( \frac{1}{2}\right)^n e_{t-n} + e_{t-1} = \\
10\left(\frac{1}{2}\right)^{t-1} + e_{t-1} + \frac{1}{2}e_{t-2} + \left(\frac{1}{2}\right)^2e_{t-3} + \dots +
  \left(\frac{1}{2}\right)^{n-1}e_{t-n} \qquad \square
\]

### b {-}

$\text{E}(Y_t) = 10 \left( \frac{1}{2}\right)^t$ varies with $t$ and thus is
not stationary.

## Stationary AR(1)

### a {-}

\[
\text{E}(W_t) = \text{E}(Y_t + c\phi^t) = \text{E}(Y_t) + \text{E}(c\phi^t) = 0 + c\phi^t = c\phi^t \tag*{$\square$}
\]


### b {-}

\[
\phi(Y_{t-1}+c\phi^{t-1})+e_t = \phi Y_{t-1} + c\phi^t + e_t = \phi \left( \frac{Y_t - e_t}{\phi}\right) + c \phi^t + e_t = Y_t + c\phi^t \tag*{$\square$}
\]

### c {-}

No, \text{E}(W_t) is not free of $t$.

## Find a simpler model (1)

This is similar to an AR(1) with $\rho_k = -(-0.5)^k$.

```{r}
ARMAacf(ar = -0.5, lag.max = 7)
ARMAacf(ma = -c(0.5, -0.25, 0.125, -0.0625, 0.03125, -0.0015625))
```

## Find a simpler model (2)

This is like an ARMA(1,1) with $\phi = -0.5$ and $\theta = 0.5$.

\[
\begin{cases}
\psi_1 = \phi - \theta = 1\\
\psi_2 = (\phi - \theta)\phi = -0.5
\end{cases}

\implies

\begin{cases}
\theta = 0.5\\
\phi = -0.5
\end{cases}
\]

```{r}
ARMAacf(ma = -c(1, -0.5, 0.25, -0.125, 0.0625, -0.03125, 0.015625))
ARMAacf(ar = -0.5, ma = -0.5, lag.max = 8)
```

## ARMA in disguise

### a {-}

\[
\begin{gathered}
\text{Cov}(Y_t, Y_{t-k}) = \text{Cov}(e_{t-1}-e_{t-2}+0.5e_{t-3}, e_{t-1-k}-e_{t-2-k}+0.5e_{t-3-k}) = \\
\gamma_k = \begin{cases}
             \sigma_e^2 + \sigma_e^2 + 0.25\sigma_e^2 = 2.25\sigma_e^2 & k = 0\\
             -\sigma_e^2-0.5\sigma_e^2=-1.5\sigma_e^2 & k = 1\\
             0.5\sigma_e^2 & k = 2
           \end{cases}
\end{gathered}
\]

### b {-}

This is an ARMA(p,q) in the sense that $p = 0$ and $q = 2$, that is, it is in
fact an MA(2) process $Y_t = e_t - e_{t-1}+0.5e_{t-2}$ with $\theta_1 = 1, \theta2 = -0.5$.

## Equivalence of statements

\[
1 - \phi_1x - \phi_2x^2 - \dots - \phi_p x^p \implies
x^k \left( \left(\frac{1}{k}\right)^p - \phi_1 \left(\frac{1}{k}\right)^{p-1} \dots - \phi_p \right)
\]

Thus if $x_1 = G$ is a root to the above, $\frac{1}{x_1} = \frac{1}{G}$ must be a root to
\[
x^p - \phi_1 x^{p-1} -\phi_2 x^{p-2} - \dots -\phi_p
\]

## Covariance of AR(1)

### a {-}

\[
\begin{gathered}
\text{Cov}(Y_t - \phi Y_{t+1}, Y_{t-k} - \phi Y_{t-k+1}) = \\
\text{Cov}(Y_t, Y_{t-k}) - \phi \text{Cov}(Y_t, Y_{t+1-k}) - \phi\text{Cov}(Y_{t+1},Y_{t-k}) + \phi^2\text{Cov}(Y_{t+1, Y_{t+1-k}}) = \\
\frac{\sigma_e^2}{1-\phi^2}\left(\phi^k - \phi \phi^{k-1} - \phi \phi^{k+1} + \phi^2 \phi\right) =\\
\frac{\sigma_e^2}{1-\phi^2}\left( \phi^k - \phi^k - \phi^{k+2} + \phi^{k+2} \right) = 0
\end{gathered}
\]

### b {-}

First,

\[
Y_{t+k} = \phi Y_{t+k-1} + e_t + k \implies
\]

\[
\begin{split}
\text{Cov}(b_t,  Y_{t+k}) & = \text{Cov}(Y_t - \phi Y_{t+1, Y_{t+n}})\\
                          & = \text{Cov}(Y_t,Y_{t+k}) - \phi \text{Cov}(Y_{t+1}, Y_{t+k)})\\
                          & = \frac{\sigma_e^2}{1-\phi^2}\phi^k - \phi \frac{\sigma_e^2}{1-\phi^2}\phi^{k-1}\\
                          & = 0 & \square
\end{split}
\]

## Recursion

### a {-}

\[
\begin{split}
E(Y_0) & = E[c_1e_0] = c_1E[e_0] = 0\\
E(Y_1) & = E(c_2Y_0 + e_1) = c_2 E(Y_0) = 0\\
E(Y_2) & = E(\phi_1 Y_1 + \phi_2 Y_0 + e_t) = 0\\
       & \vdots \\
E(Y_t) & = E(\phi_1 Y_{t-1} + \phi_2 Y_{t-2}) = 0
\end{split}
\]

### b {-}

We have
\[
\begin{cases}
\text{Var}(Y_0) = c_1^2\text{Var}(e_0) = c_1^2\sigma_e^2\\
\text{Var}(Y_1) = c_2^2\text{Var}(Y_0) + \text{Var}(e_0) = c_2^2c_1^2\sigma_e^2=\sigma_e^2(1+c_1^2c_2^2)
\end{cases}  \implies
\]

\[
\begin{cases}
c_1^2\sigma_e^2 = \sigma_e^2(1+c_1^2c_2^2) & \iff c_1^2(1-c_2^2) = 1\\
c_1^2 = \frac{1}{1-c_2^2} & \iff c_1 = \sqrt{\frac{1}{1-c_1^2}} = \frac{1}{\sqrt{1-c_1^2}}
\end{cases}
\]
with covariance
\[
\begin{gathered}
\text{Cov}(Y_0, Y_1) = \text{Cov}(c_1e_0,c_2 c_1 e_0 + e_1) = \text{Cov}(c_1e_0,c_2 c_1 e_0) + \text{Cov}(c_1e_0, e_1) =\\
c_1^2c_2\sigma_e^2 + c_1\text{Cov}(e_0, e_1) = c_1^2c_2\sigma_e^2 + 0
\end{gathered}
\]
and autocorrelation
\[
\rho_1 = \frac{c_1^2c_2\sigma_e^2}{\sqrt{(c_1^2)^2}} = c_2 
\]
so we must choose
\[
\begin{cases}
c_2 = \frac{\phi_1}{1-\phi_2} \\
c_1 = \frac{1}{\sqrt{1-c_2^2}}
\end{cases}.
\]

### c {-}

The process can be transformed by scaling and standardizing it and then shifting
with any given mean.
\[
\frac{Y_t \sqrt{y_0}}{c_1} + \mu
\]

## Final exercise

### a {-}

\[
\begin{split}
Y_t & = \phi Y_{t-1} + e_t\\
    & = \phi(\phi Y_{t-2}+e_{t-1}) + e_t = \phi^2 Y_{t-2} + \phi e_{t-1} + e_t\\
    & \vdots \\
    & = \phi^t Y_{t-t} + \phi e_{t-1} + \phi^2 e_{t-2} + \dots + \phi^{t-1}e_1+e_t & \square
\end{split}
\]

### b {-}

\[
E(Y_t) = E(\phi^t Y_0 +\phi e_{t-1} + \phi^2 e_{t-2} + \dots + \phi^{t-1}e_1+e_t) =\phi^t\mu_0
\]

### c {-}

\[
\begin{split}
\text{Var}(Y_t) & = \text{Var}(\phi^t Y_0 + \phi e_{t-1}+\phi^2e_{t-2} + \dots + \phi^{t-1}e_1)\\
                & = \phi^{2t}\sigma_0^2+\sigma_e^2 \sum_{k=0}^{t-1}(\phi^2)^k\\
                & = \sigma_e^2 \frac{1-\phi^{2n}}{1-\phi^2} + \phi^{2t}\sigma_0^2 \quad\text{if $\phi \neq 1$ else}\\
                & = \text{Var}(Y_0) + \sigma_e^2 t = \sigma_0^2 + \sigma_e^2 t
\end{split}
\]

### d {-}

If $\mu_0 = 0$ then $E(Y_t) = 0$ but for $\text{Var}(Y_t)$ to be free of t, 
$\phi$ cannot be 1.

### e {-}

\[
\text{Var}(Y_t) = \phi^2 \text{Var}(Y_{t-1}) + \sigma_e^2 \implies \phi^2 \text{Var}(Y_t) + \sigma_e^2
\]
and
\[
\text{Var}(Y_{t-1}) = \text{Var}(Y_t)(1-\phi^2) = \sigma_e^2 \implies \text{Var}(Y_t) = \frac{\sigma_e^2}{1-\phi}
\]

and then wew must have $|\phi| < 1$.
